---
title: "Biostat 203B Homework 3"
author: "Tomoki Okuno"
subtitle: Due Feb 25 @ 11:59PM
output:
  html_document: 
    toc: yes
    toc_depth: 4
  # ioslides_presentation: default
---

Display machine information:
```{r}
sessionInfo()
```

Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(dbplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
library(DBI)
library(bigrquery)
```

## Q1. Compile the ICU cohort in HW2 from the Google BigQuery database 

Below is an outline of steps.

1. Load the service account token.
```{r}
# path to the service account token 
if (Sys.info()[["nodename"]] == "biostat-203b-teaching-server") {
  # on teaching server
  satoken <- "/mnt/mimiciv/1.0/biostat-203b-2022winter-3fdc2392ac39.json"
} else {
  # on my own Mac laptop
  satoken <- "/Users/huazhou/Documents/github.com/ucla-biostat-203b/2022winter/hw/hw3/biostat-203b-2022winter-3fdc2392ac39.json"
}
# BigQuery authentication using service account
bq_auth(
  path = satoken,
  email = "mimiciv-bigquery@biostat-203b-2022winter.iam.gserviceaccount.com",
  scopes = c("https://www.googleapis.com/auth/bigquery",
     "https://www.googleapis.com/auth/cloud-platform")
)
```

1. Connect to BigQuery database `mimic4_v1_0_203b` in GCP (Google Cloud Platform), using the billing account `biostat-203b-2022winter`.
```{r}
# Connect to the BigQuery database `biostat-203b-2022winter.mimic4_v1_0_203b`
con <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2022winter",
    dataset = "mimic4_v1_0_203b",
    billing = "biostat-203b-2022winter"
)
con
```

2. List all tables in the `mimic4_v1_0_203b` database.
```{r}
dbListTables(con)
```

3. Connect to the `icustays` table.  
```{r}
# full ICU stays table
icustays_tble <- tbl(con, "icustays") %>%
  show_query() %>%
  print(width = Inf)
```
We only keep the first ICU stay. Following code is kind of a hack, using the `summarise_all(min)` function. It seems that `slice_min()`, `slice_head()`, `distinct(, .keep_all = TRUE)` don't work with `dbplyr`+`bigrquery` at the moment.  
```{r}
icustays_tble <- icustays_tble %>% 
  select(subject_id, intime) %>% 
  group_by(subject_id) %>% 
  summarise_all(min) %>%
  left_join(icustays_tble, by = c("subject_id", "intime")) %>%
  show_query() %>%
  print(width = Inf)
```


4. Connect to the `admissions` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id` and `hadm_id`).

**Solution:** Here is my code. 
```{r warning = FALSE}
# import admissions table and only keep the patients for the first ICU stay
admissions_tble <- tbl(con, "admissions") %>%
  semi_join(icustays_tble, by = c("subject_id", "hadm_id")) %>%
  show_query() %>%
  print(width = Inf)
```

5. Connect to the `patients` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

**Solution:** Here is my code. 
```{r warning = FALSE}
# import patients table and only keep the patients for the first ICU stay
patients_tble <- tbl(con, "patients") %>%
  semi_join(icustays_tble, by = c("subject_id")) %>%
  show_query() %>%
  print(width = Inf)
```

6. Connect to the `labevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the lab items listed in HW2.

**Solution:** Here is my code. I ended up not using labels on app (inputting them directly in `ui`).
```{r warning = FALSE}
# full labevents table
labevents_tble <- tbl(con, "labevents") %>%
  show_query() %>%
  print(width = Inf)

d_labitems_tble <- tbl(con, "d_labitems") %>%
  show_query() %>%
  print(width = Inf)

choice_lab <- c(50912, 50971, 50983, 50902, 50882, 
                51221, 51301, 50931, 50960, 50893)

labevents_tble <- labevents_tble %>%
  # filter out 10 lab items
  select(subject_id, itemid, charttime, valuenum) %>%
  filter(itemid %in% choice_lab) %>%
  semi_join(icustays_tble, by = c("subject_id")) %>%
  show_query() %>%
  print(width = Inf)
```
Only keep the first lab measurements during ICU stay and pivot lab items to become variables/columns.

**Solution:** Here is my code. I avoided an error in BigQuery by adding `lab` at the beginning of the `itemid`, using `names_prefix`.
```{r warning = FALSE}
labevents_tble <- labevents_tble %>%
  # keep only lab measurements between intime and outtime
  left_join(select(icustays_tble, subject_id, intime, outtime),
            by = c("subject_id")) %>%
  filter(charttime >= intime, charttime <= outtime) %>%
  show_query() %>%
  print(width = Inf)

labevents_tble <- labevents_tble %>%
  # sort charttime in ascending order by subject_id and item_id
  select(subject_id, itemid, charttime) %>%
  group_by(subject_id, itemid) %>%
  summarise_all(min) %>%
  left_join(labevents_tble, by = c("subject_id", "itemid", "charttime")) %>%
  # keep only 11 columns and spread itemid and valuenum
  select(-charttime, -intime, -outtime) %>%
  # spread valuenum by itemid 
  pivot_wider(names_from = itemid, 
              values_from = valuenum, 
              names_prefix = "lab") %>%
  show_query() %>%
  print(width = Inf) 
```

7. Connect to `chartevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the chart events listed in HW2.

**Solution:** Here is my code, based on the same idea as Q1-6.
```{r warning = FALSE}
# full chartevents table
chartevents_tble <- tbl(con, "chartevents") %>%
  show_query() %>%
  print(width = Inf)

d_items_tble <- tbl(con, "d_items") %>%
  show_query() %>%
  print(width = Inf)

choice_chart <- c(220045, 220181, 220179, 223761, 220210)

chartevents_tble <- chartevents_tble %>%
  # only containing these items for the patients in icustays_tble
  select(stay_id, itemid, charttime, valuenum) %>%
  # only containing these items for the patients in icustays_tble
  filter(itemid %in% choice_chart) %>%
  semi_join(icustays_tble, by = c("stay_id")) %>%
  # # add label in d_items
  # left_join(select(d_items_tble, itemid, label), by = c("itemid")) %>%
  show_query() %>%
  print(width = Inf)
```
Only keep the first chart events during ICU stay and pivot chart events to become variables/columns.

**Solution:** Here is my code, based on the same idea as Q1-6.
```{r warning = FALSE}
chartevents_tble <- chartevents_tble %>%
  # add intime & outtime from icustays_tble to find the first vital measurement
  left_join(select(icustays_tble, stay_id, intime, outtime), 
            by = c("stay_id")) %>%
  # keep only vital measurements between intime and outtime
  filter(charttime >= intime, charttime <= outtime) %>%
  show_query() %>%
  print(width = Inf)

chartevents_tble <- chartevents_tble %>%
  select(stay_id, itemid, charttime) %>%
  # sort charttime in ascending order by subject_id and item_id
  group_by(stay_id, itemid) %>%
  # keep only the first vital measurement by group
  summarise_all(min) %>%
  left_join(chartevents_tble, by = c("stay_id", "itemid", "charttime")) %>%
  # restrict columns and spread label and valuenum
  select(-charttime, -intime, -outtime) %>%
  # spread valuenum by itemid 
  pivot_wider(names_from = itemid, 
              values_from = valuenum, 
              names_prefix = "chart") %>%
  show_query() %>%
  print(width = Inf)
```

8. Put things together. This step is similar to Q7 of HW2. Using one chain of pipes `%>%` to perform following data wrangling steps: (i) start with the `icustays_tble` for the first ICU stay of each unique patient, (ii) merge in admissions and patients tables, (iii) keep adults only (age at admission >= 18), (iv) merge in the labevents and chartevents tables, (v) create an indicator for 30-day mortality, (vi) save the final tibble to an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.

**Solution:** Here is my code. The difference from HW2 is:

- age at admission >= 18 instead of >18
- using `dod` instead of `deathtime` due to limitations in BigQuery
```{r warning = FALSE}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
# (i) start with the `icustays_tble` for the first ICU stay
icu_cohort <- icustays_tble %>%
  # (ii) merge in admissions and patients tables
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) %>%
  left_join(patients_tble, by = c("subject_id")) %>%
  # (iii) keep adults only (age at admission >= 18)
  mutate(age_hadm = anchor_age + year(admittime) - anchor_year) %>%
  filter(age_hadm >= 18) %>%
  # (iv) merge in the labevents and chartevents tables
  left_join(labevents_tble, by = c("subject_id")) %>%
  left_join(chartevents_tble, by = c("stay_id")) %>%
  # (v) create an indicator for 30-day mortality
  mutate(thirty_day_mort = 
           ifelse(is.na(dod), FALSE,
                  DATE_DIFF(date(dod), date(admittime), DAY) <= 30)) %>%
  # (vi) save the final tibble to an `icu_cohort.rds` in `mimiciv_shiny`
  collect() %>%
  write_rds("mimiciv_shiny/icu_cohort.rds") %>%
  # show_query() %>%
  print(width = Inf)

# check the number of patients who died within 30 days
icu_cohort %>%
  select(admittime, dod, thirty_day_mort) %>%
  filter(thirty_day_mort == TRUE) %>%
  summarise(n = n()) %>%
  print(width = Inf)
```
**Comment:** I got 5,436 patients who died withn 30 days. If changing `age_hadm >= 18` to `> 18`, the number will be 5,435.

Close database connection and clear workspace.
```{r, eval=F}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should provide easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort.

**Solution:** My app (app.R) was saved in the `mimiciv_shiny` folder. I made two types of summaries: using 30-day mortality as an indicator (like HW2) and not. For numerical variables, I removed extreme values to make the graphs readable.
